---
editor_options:
  markdown:
    wrap: 72
---

# MACHINE LEARNING IN HEALTHCARE

# PART 1 - CLUSTERING ASSIGNMENT

DATASET - "heart-c.csv" (Clustering Data)

**Question 1:\
Run K-means clustering on the above heart disease dataset and answer the
following questions**

K-means is a form of unsupervised Machine Learning algorithm which is
used for clustering data points into distinct groups where k is the
number of clusters. Global Attributes like k is the number of clusters,
maximum number of iterations, random seed which ensures reproducibility
of results Firstly, k-means algorithm allocates k number of cluster
centroids n the data space and are optimised by decreasing the
metric-"goodness of fit" this metric is the sum of squared distances
between each data point and its corresponding cluster centroid. The
algorithm iteratively reassigns data points to the closest centroid and
the centroids positions are updated by the new data points which is
assigned. This iteration is done until there is no change in the
centroids or until the maximum iterations is reached. Here, it has found
the best centroid positions and clusters are stable. this is called the
Convergence point, where a data point is assigned to a closest cluster
centroid, is the final cluster for each of the data point. When it comes
to deciding the number of clusters that is K value, it requires
interpretation, validation and observation. So that the clusters makes
sense and aligns with real-world patterns.

To perform K-means clustering on the data provided, importing the data
is the first step. The data is in CSV file, which is in structured
format.

#### importing R packages #collection of R packages(ggplot2,dplyr,tidyr) #data manipulation and visualization

```{r}
library(tidyverse)  
library(factoextra) #this package is used for creating plots and visualization
library(cowplot)    #creating plots and graphs
library(cluster)    #clustering functions
```

#Importing the heart disease data set

```{r}
heart_disease <- read_csv("/Users/neha-/Desktop/PMIM402J-MLH/clustering data.csv")
```

since the data set is in structured csv format, I am using read.csv()
function the data is read into the data frame named 'heart_disease'

```{r}
heart_disease <- na.omit(heart_disease)
```

removing rows with missing data Using na.omit() function to remove rows
from data frame which might contain missing values like NA, which will
ensure we have complete data further. Fixing the global seed for
reproducibility(across different runs of the script)

```{r}
set.seed(15) 
```

to set a seed for the random number generator. This will help in getting
consistent results which makes it easier to compare and validate the
output. Since I will be creating plots, I am setting the options for
size of plots which will be created subsequently in this current R
session.

```{r}
options(repr.plot.width=14, repr.plot.height=10)
```

the size of plots will have width of 12 inches and height of 8 inches.
Explicitly setting ggplot2 theme to default cowplot theme

```{r}
theme_set(theme_cowplot())
```

1\. Why should the attribute "class" in heart-c.csv("num")not be
included for clustering?

In the d ataset, the attribute 'num' represents Diagnosis of heart
disease.Basically it is classified attribute, and is not needed in
Clustering Because Clustering will be done without using ground truth
labelled information By removing 'diagnosis' i.e., 'num' attribute, The
Clustering algorithm is executed only on other attributes like age,
sex,..which helps in identifying natural groupings or clusters based on
similarities. renaming attributes to make it easier to understand

Saving a copy of the original data set with the ground truth 'num'
attribute for later, so I can compare it with clustering results.

```{r}
heart_original <- heart_disease

```

Removing the diagnosis attribute so that it is suitable for Unsupervised
Clustering Process

```{r}
heart_disease <- subset(heart_disease, select = -c(num))

#Viewing the dataset after removing the 'diagnosis' attribute.

head(heart_disease, 20)

```

By understanding the nature and characteristics of each attribute, I see
that the data set comprises of both continuous and categorical
attributes, but K-means Clustering is suitable for Continuous attributes
because it requires projecting data into a Continuous domain. Continuous
Attibutes are numerical, by taking real value and represents
measurements on continuous scale range.

Continuous attributes: age,
resting_blood_pressure,cholesterol,max_heart_rate,
relative_st_depression, and fluoroscopy_count. Categorical attributes:
sex, chest_pain_type, fasting_blood_sugar, resting_ecg_result,
has_exercise_angina, peak_st_slope, thallium_test_result, diagnosis

Since the projection of categorical attributes is meaningless, I am
filteringonly continuous attributes

```{r}
heart_continuous <- Filter(is.numeric, heart_disease) # is.numeric function checks if variable is numeric - True or not - False. # Filter() function is used to select numeric columns 
```

\# Viewing the Continuous attributes only

```{r}
 head(heart_continuous)
```

Applying standardization or z-score normalization which transforms in a
way that mean becomes 0 and standard deviation becomes 1

```{r}
heart_continuous <- scale(heart_continuous) # scale() function is used to standardize columns of data frame.
```

Viewing the heart disease data set with only continuous attributes

```{r}
head(heart_continuous)
```

2.Run the K-means algorithm and provide reasoning for the optimum value
of K. I can now apply k-means Clustering analysis, but for that first,
optimal value for k which produces best clustering is required. I am
using few statistical methods, to compare and find the effectiveness of
different k values. By calculating within-cluster sum of squares (WCSS).
WCSS is sum of squared distances between data points and their assigned
cluster centroids which are compact and tightly grouped clusters.

This is called the Elbow Method, the name comes from shape of the plot,
where line forms an 'elbow-like' bend at the optimal value of k.

Plotting the elbow curve using 'fviz_nbclust()' function

```{r}
fviz_nbclust(
  x = heart_continuous, #preprocessed dataset with only continuous attributes
  FUN = kmeans,    # k means clustering function is used
  method = "wss", # Within Sum of Squares method - calculates WCSS for k value
  k.max = 10, # maximum value of K:10, all the values from 1 to 10 is plotted
  linecolor = "black"
) +
  theme(
    panel.grid.major = element_line() # Adds gridlines to the plot for clarity
  )
```

3.Which features would you expect to be less useful when using K-means
and why? From the plot, there's no obvious elbow point and few more
comparisons are needed before deciding the best k value. when k = 1, a
single cluster is formed which is not informative and it groups all data
together without any differentiation when k \>=6, as number of clusters
increases beyond 5 or 6 the decrease in WCSS, when k = 2\~5, curve shows
more gradual decrease in WCSS and no clear elbow point the levelling off
indicates the clustering with 2 to 5 clusters might be reasonable
additional techniques like Silhouette analysis, Principal Component
Analysis(PCA) and Gap Statistics can be applied.

Gap Statistics compare the clustering performance to that of a random
distribution, Maximizing it to find the optimal value .This compares the
WCSS of actual clustering to the reference distribution to estimate gap
between clustering performance and random data.

Plotting gap statistics

```{r}
fviz_nbclust(
  x = heart_continuous,
  FUN = kmeans,    # k means type of function
  method = "gap_stat", # gap statistics method
  nboot = 28,    # Number of start permutations
  k.max = 10,
  linecolor = "black"
) +
  theme(
    panel.grid.major = element_line()    # Adding gridlines to the plot
  )
```

when𝑘=2, has the largest gap statistic and is looking to be the most
optimal

clustering.

Performing Principal Component Analysis(PCA) for visualising cluster in
2D

generating k means clustering models with k in range [2,k]

we have now understood that k=1 creates a meaningless cluster

```{r}
kmeans_model <- function(dataset, k_max , n_start = 25) {
  # Store results in a list
  results <- c()
  
  # Iterate through range [2, k_max]
  for (k in 2:k_max) {
    # Assign results[k] to k-means
    results[[as.character(k)]] <- kmeans(
      x = dataset,
      centers = k,
      nstart = n_start,    # Trying different random starting positions and 
                           # picking up the best one
      iter.max = n_start   # Increase iteration limiter
    )
  }
  return(results)
}
```

I am using kmeans_model function to generate k means clustering models
for k in range of 2 to 10 and print description of one of them.

Maximum value of k to consider

```{r}
k_max = 10
```

Calculate and storing the results of k-means run from 2 \~ k_max

```{r}
kmeans_results <- kmeans_model(heart_continuous, k_max)
```

Preview a clustering result

```{r}
kmeans_results[[2]]
```

Plotting using PCA for all k values. PCA is a reduction technique which
helps in visualising high dimensional data in a lower dimensional space
like 2D or 3D.

**Increase plot size**

```{r}
options(repr.plot.width=28, repr.plot.height=20)
```

#### Create chart objects for each clustering model

```{r}
chart_object <- list()
```

the below loop iterates over the results of kmeans clustering for
different k

values from 2 to length of kmeans_results

```{r}
for (i in 2:length(kmeans_results)) {
    # Visualise each clustering result
    chart_object[[i]] <- fviz_cluster(
        kmeans_results[[i]],
        data = heart_continuous,
        geom = "point",
        show.clust.cent = TRUE
    ) + 
    ggtitle(
        paste(c("k =", i, "\navg cluster size =", round(mean(kmeans_results[[i]]
                                                             $size),1),"points"),
              collapse = " ")    # Title each chart: "k = i; avg cluster size = 
                                 # n points"
    )
}
```

# View the charts

when k=2, the plots are very clear and gives clarity, it is segmented

For all other vlues of k, it has at least some overlap. for the purpose
of evaluating, I am now including the data set with ground truth
labelled data

When k =2, an increased level of heart disease are colored green and
those without are colored red

#The cluster centroids and outline are in Teal and Magenta for cluster 1
and 2.

Attach ground truth attribute to k=2 chart color channel and visualise

#From this chart object, it looks like the clustering is largely
accurate, #with green and red points within cluster 1 and 2 #For a more
quantitative analysis of the clustering's performance, we'll #determine
its sensitivity, specificity, and accuracy.

We need to get the clustering result for each sample, which we can most
easily, obtain by extracting it from the above chart object

Extracting chart object data from k=2 plot

```{r}
k2_data <- layer_data(chart_object[[2]], 1)
```

Extracting cluster attribute from chart object and converging it with
the original heart data table

```{r}
heart_original$cluster <- k2_data$group
```

Observe new cluster attribute

```{r}
head(heart_original, 5)
```

I am able to directly compare the results of the clustering with the
ground truth labels, assuming that patients with an increased level of
heart disease where diagnosis = "\>50_1" are in cluster 2, and those
without where diagnosis = "\<50" are in cluster 1.

```{r}
pos = c() 
neg = c()
```

#### True positives

```{r}
pos["True"] = nrow(heart_original[which(heart_original$num == ">50_1"
                                             & heart_original$cluster == 2),])
```

#### False positives

```{r}
pos["False"] = nrow(heart_original[which(heart_original$num == ">50_1"  & heart_original$cluster == 1),])
```

#### True negatives

```{r}
neg["True"] = nrow(heart_original[which(heart_original$num == "<50"  & heart_original$cluster == 1),])
```

#### False negatives

```{r}
neg["False"] = nrow(heart_original[which(heart_original$num == "<50"  & heart_original$cluster == 2),])
```

#### Constructing confusion matrix

```{r}
confusion_matrix = data.frame(pos, neg)
```

#### Viewing confusion matrix

```{r}
print(confusion_matrix)
```

#### To evaluate the clustering performance, I am calculating the metrics:

```{r}
paste("Sensitivity:", round(pos["True"] / (pos["True"] + pos["False"]), 3))
paste("Specificity:", round(neg["True"] / (neg["True"] + neg["False"]), 3))
paste("Accuracy:", round((pos["True"] + neg["True"]) / (pos["True"] + 
                                                          neg["True"] + 
                                                          pos["False"] + 
                                                          neg["False"]),3))
```

Question 2:

Run the hierarchical clustering on above heart disease dataset, and
answer the following questions

Hierarchical clustering techniques is unsupervised machine learning
technique, it groups similar data points into clusters based on their
similarity and dissimilarity. Each data point starts as its own and
builds a tree like hierarchical structure of clusters, forming a tree
called as Dendrogram.

This gives more flexible, clear and accurate explanation than k-means
because it doesn't require to specify the number of clusters beforehand.
Agglomerative hierarchical cluster starts with each data point as its
own cluster and iteratively merges most similar clusters until all data
points are combined into one cluster. ( bottom-up )

Divisive hierarchical cluster starts with all data points in one cluster
and recursively divides the cluster into smaller clusters until each
data point is in its own individual cluster. ( top-down)

#Divisive clustering aims to achieve fewer, larger clusters.

# Preview

```{r}
head(heart_continuous)
```

1.Show the clustering results in a tree structure and provide reasoning
for the optimal number of clusters

#We perform the divisive clustering using Euclidean distances(measures
straight line distance between two points in multi dimensional space) as
our dissimilarity measure and represent the splitting of our data set
using a dendrogram

#diana() function is a part of cluster package which allows to perform
divisive \# hierarchical clustering

```{r}
divisive <- diana(heart_continuous, metric = "euclidean")

fviz_dend(
  divisive,
  main = "Divisive hierarchical clustering of the heart data set (continuous vars only)",
  cex = 0.4,    # Resize leaf node labels
) + scale_y_continuous(
  name = "Dendrogram height",
  n.breaks = 10    # Individual y-axis labels
)
```

#The dendrogram allows to see which groups that samples have been
divided into. Reading from top to bottom, the divisive clustering has
sectioned the data until each sample is in its own cluster.

```{r}
k = 5
```

```{r}
fviz_dend(
  divisive,
  cex = 0.4,
  main = "Divisive hierarchical clustering of the heart data set (continuous 
  vars only) - clusters highlighted",
  k = k    # Choose split point
) + geom_hline(
  yintercept = 9,    # Should be (height - k) for a dendrogram that isn't squished
  linetype = "dashed",
  color = "navy"
) + scale_y_continuous(
  name = "Dendrogram height",
  n.breaks = 10    # Individual y-axis labels
)
```

#The dendrogram is cluttered a little and y-intercept line is a bit off,
however the different colours show the split of clusters clearly. In
this case, the red cluster is far bigger than the other clusters ---
indicating that the divisive clustering is kind of unbalanced.

In summary, the divisive coefficient is efficient to measure an overall
'goodness of fit' of the clustering based on how well separated and
balanced each cluster is --- values are in the range [0, 1], with a
value of 1 meaning clusters are completely separated and each cluster is
the same size.

```{r}
paste("Divisive coefficient:", round(divisive$dc, 5))
```

\#'Divisive coefficient: 0.84506'

#which is a very high coefficient value, makes me believe that divisive
clustering is effective at producing clusters; however it is difficult
to take a single metric like this with no context or comparison and
derive a conclusion. Additionally, there are certain aspects that the
coefficient cannot tell,like how many clusters are most optimal or by
which factors the data set has been clustered by --- further analysis is
required for this.

Comparing with agglomerative clustering - Using differing linkage
methods, I will compare the results of Divisive clustering and
Agglomerative clustering which should be able to give ' goodness of fit'
for each dendrogram. Very similarly to

2.Describe the link method you used.

3.What are the strengths and limitations of this link method in
hierarchical clustering?

#The five different linkage methods which will be evaluated are:
1.single-linkage, where the closest single pair of elements are not in
the same cluster are considered as linkage criteria 2.complete-linkage,
where the furthest away single pair of elements not in the same cluster
are considered as linkage criteria 3.weighted average-linkage (WPGMA),
where the linkage criteria is a simple average of the distance between
clusters 4.unweighted average-linkage (UPGMA), where the linkage
criteria is the average distance between clusters, where the number of
samples in the cluster are accounted for - making it 'unweighted'
5.Ward's method, where the linkage criteria is the pair of clusters
which, after merging, increases the total within-cluster variance by the
smallest amount; minimising the within-cluster variance

# Calculating Euclidean dissimilarity matrix

```{r}
diss_matrix = dist(heart_continuous, method = "euclidean")

aggl_single   <- agnes(diss_matrix, method = "single")
aggl_complete <- agnes(diss_matrix, method = "complete")
aggl_wpgma    <- agnes(diss_matrix, method = "weighted")
aggl_upgma    <- agnes(diss_matrix, method = "average")
aggl_ward     <- agnes(diss_matrix, method = "ward")

aggl = list(
  "single" = aggl_single,
  "complete" = aggl_complete,
  "WPGMA" = aggl_wpgma,
  "UPGMA" = aggl_upgma,
  "Ward's method" = aggl_ward
)
```

# Create chart objects for each clustering model

```{r}
chart_object <- list()

for (i in 1:length(aggl)) {
  # Visualise each clustering result
  chart_object[[i]] <- fviz_dend(
    aggl[[i]],
    cex = 0.4
  ) + 
    ggtitle(
      # Dynamically title each chart
      paste(c(
        "Agglomerative clustering using the", names(aggl)[[i]], "linkage method\n",
        "Agglomerative coefficient:", round(aggl[[i]]$ac, 3)
      ), collapse = " ")
    ) +
    scale_y_continuous(
      name = "Dendrogram height",
      n.breaks = 10    # Individual y-axis labels
    )
}
```

# Increasing the plot size again

```{r}
options(repr.plot.width=42, repr.plot.height=24)
```

# View the charts

```{r}
plot_grid(plotlist = chart_object)
```

#From the comparison that Ward's method is, at least in terms of
clustering quality, the superior linkage method --- with an excellent
agglomerative coefficient of 0.959 and visually balanced and
well-separated. Conversely, single-linkage and UPGMA have the lowest
coefficient scores and visually the dendrograms are tightly packed,
skewed, and overall poorly separated. It is worth reiterating the fact
that our 'best' clustering method is meaningless if it is not applied to
a particular issue and analysed by a domain expert; the coefficient
score simply ranks the structure of the dendrogram rather than
evaluating the accuracy of its clustering.

'Ward's method' of selecting linkage criteria produces tight, packed
clusters that is good at determine large-scale patterns in the data.
